### 2013-03-24 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating disseration chapter structure

%%---------------------------------------------------------------------------%%
\section{Domain-to-Domain Communication}
\label{sec:mc_comm_costs}
With Wagner's work, we observed that domain overlap implemented with a
domain decomposed Monte Carlo setting reduces the amount of nearest
neighbor communication that must occur during transport. This
communication, expected to be expensive, was analyzed by Siegel and
colleagues for idealized, load balanced situations without domain
overlap \citep{siegel_analysis_2012}. They arrive at a few empirical
relationships and generate a numerical experiment that provides
insight into the feasibility of large scale domain decomposed Monte
Carlo as well as the relationship of communication to system
parameters. In their work, they define a simple 3-dimensional system
subdivded into Cartesian block domains with each processor containing
a single domain. Given this geometry, a particle can either be
absorbed in its current domain or be transported to one of the 6
adjacent domains owned by another process.

To quantify the number of particles that leak out of the local domain
they define a leakage fraction, $\lambda$, as:
\begin{equation}
  \lambda =
  \frac{average\ \#\ of\ particles\ leaving\ local\ domain}{total\ 
    of\ \#\ of\ particles\ starting\ in\ local\ domain}\:.
  \label{eq:leakage_fraction}
\end{equation}
For their studies, they assumed that the value of $\lambda$ was
dependent on the total cross section of the system and the same for
all processes for simplicity and load balancing. Using this
relationship and their simple geometry, they were able to relate the
amount of compute time required to transport all source particles to
the number of source particles, the leakage fraction, the latency and
bandwidth of communication, the ratio of local to global geometry
size, and the total cross section of the system.

What they found was that there was indeed an optimum ratio of local
domain to global domain size that results in a minimum total compute
time. They also found that this ratio was only dependent on the
latency and bandwidth constraints of the parallel machine. We expect
this because nearest neighbor communication patterns scale with the
volume and surface sizes of the local domains. The larger the volume,
the more compute time is spent within the local volume and therefore
at some point you are limited by the latency of the system, or the
time it takes to receive the data to operate on. However, at a certain
point, when the surface of those volumes becomes large enough, the
amount of data that must be communicated grows large enough that it
may overcome the bandwidth limitations of the system, resulting in a
decrease in performance. In their numerical experiments, they used the
above geometry and leakage rates based on that computed for a reactor
system for a single material problem with a total cross section
equivalent to that of homogenized light water reactor. Given the
optimal local domain size, they observed for load balanced situations
that nearest neighbor communication costs during transport are a small
fraction of total compute time, leading to the conclusion that domain
decomposition is feasible for large-scale Monte Carlo calculations.

\subsection{Load Balancing Concerns}
\label{subsec:mc_load_balancing}
Although domain decomposition was shown to be efficient in a perfectly
load balanced situation in Siegel's work \citep{siegel_analysis_2012},
careful consideration must be made for situations where this is not
the case. Given the stochastic nature of the problem and lack of a
globally homogeneous domain, parallel Monte Carlo simulations are
inherently load imbalanced. Procassini and others worked to alleviate
some of the load imbalances that are generated by both particle and
spatial parallelism and are therefore applicable to the MSOD algorithm
\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
in which the number of times a particular domain was replicated was
dependent on the amount of work in that domain (i.e. domains with a
high particle flux and therefore more particle histories to compute
require more work). In this variation, domains that require more work
will be replicated more frequently at reduced particle counts in each
replication. Furthermore, Procassini and colleagues noted that as the
simulation progressed and particles were transported throughout the
domain, the amount of replication for each domain would vary as
particle histories began to diffuse, causing some regions to have
higher work loads and some to have smaller work loads than the initial
conditions.

Consider the example in Figure~\ref{fig:procassini_example} adapted
from Procassini's work.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/mc_background/procassini_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Example illustrating how domain decomposition can
      create load balance issues in Monte Carlo.}  \textit{A domain is
      decomposed into 4 zones on 8 processors with a point source in
      the lower left zone. As the particles diffuse from the source in
      the random walk sequence as shown in the top row, their tracks
      populate the entire domain. As given in the bottom row, as the
      global percentage of particles increases in a zone, that zone's
      replication count is increased.}}
  \label{fig:procassini_example}
\end{figure}
In this example, a geometry is decomposed into 4 domains on 8
processors with a point source in the bottom left domain. To begin,
because the point source is concentrated in one domain, that domain is
replicated 5 times such the amount of work it has to do per processor
is roughly balanced with the others. As the particles begin to diffuse
away from the point source, the amount of replication is adjusted to
maintain load balance. Near the end of the simulation, the diffusion
of particles is enough that all domains have equal replication.  By
doing this, load balance is improved as each domain has approximately
equal work although each domain may represent a different spatial
location and have a differing number of histories to
compute. Compared to Wagner's work where the fission source was
distributed relatively evenly throughout the domain, fixed source
problems (and especially those that have a point-like source) like
those presented in Procassini's work will be more prone to changing
load balance requirements.

\subsection{Reproducible Domain Decomposed Results}
\label{subsec:reproducible_mc}
The 2006 work of Brunner is notable in that the Monte Carlo codes used
to implement and test the algorithms adhered to a strict policy of
generating identical results independent of domain decomposition or
domain replication as derived from the work of Gentile and colleagues
\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
given for obtaining results reproducible to machine precision for an
arbitrary number of processors and domains. Differences can arise from
using a different random number sequence in each domain and performing
a sequence of floating point operations on identical data in a
different order, leading to variations in round-off error and
ultimately a non-identical answer. They use a simple example,
recreated below in Figure~\ref{fig:gentile_example}, that illustrates
these issues.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/mc_background/gentile_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Gentile's example illustrating how domain
      decomposition can create reproducibility issues in Monte Carlo.}
    \textit{Both particles A and B start in zone 1 on processor
      1. Particle A moves to zone 2 on processor 2 and scatters back
      to zone 1 while B scatters in zone 1 and remains there. A1 and
      A2 denote the track of particle A that is in zone 1 while B1 and
      B2 denote the track of particle B that is in zone 1.}}
  \label{fig:gentile_example}
\end{figure}
In this example, the domain is decomposed on two processors with each
processor owning one of the two zones. Starting with particle A, it is
born in zone 1 and is transported to zone 2 where a scattering event
occurs. Concerning the first reproducibility issue, if the same
sequence of random numbers is not used to compute the trajectory from
the new scattering event, we cannot expect to achieve the same result
if the domain were not decomposed. If the numbers are different, the
scattering event in zone 2 may keep the particle there or even eject
it from the domain if the sequence were different. The second issue is
demonstrated by adding another particle B that remains in the
domain. In this case, an efficient algorithm will transport particle A
on processor 1 until it leaves zone 1 and then transport particle
B. Particle A will not renter the domain until it has been
communicated to processor 2, processor 2 performs the transport, and
it is communicated back to processor 1. If we are doing a track-length
tally in zone 1, then we sum the tracks lengths observed in that
zone. In the single processor, single zone case particle A would be
transported in its entirety and then particle B transported. This
would result in a tally sum with the following order of operations:
$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
of floating point operations, we cannot expect these to have an
identical result to machine precision as round-off approximations will
differ resulting in non-commutative addition.

@@

\section{Parallel Adjoint Method}
\label{sec:parallel_adjoint}
We can take much of what was learned from the survey of parallel Monte
Carlo methods for radiation transport and directly apply it to a
parallel formulation of our stochastic linear solvers. Direct analogs
can be derived from these works by noting that the primary difference
between solving a linear system with Monte Carlo methods and fixed
source Monte Carlo transport problems is the content of the Markov
chains that are generated. The transitions represented by these chains
are bound by probabilities and weights and are initiated by the
sampling of a source. In the context of transport problems, those
transitions represent events such as particle scattering and
absorption with probabilities that are determined by physical data in
the form of cross sections. For stochastic matrix inversion, those
transitions represent moving between the equations of the linear
system (and therefore the physical domain which they represent) and
their probabilities are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains. Therefore, parallel methods for Monte Carlo radiation
transport can be abstracted and we can use those concepts that apply
to matrix inversion methods as an initial means of developing a
parallel Neumann-Ulam-type solver. Based on the results observed in
the last decade of parallel Monte Carlo development, we can generate
many practical questions that this type of work could answer.

@@

\section{Parallel MCSA}
\label{sec:parallel_mcsa}
With a parallel adjoint Neumann-Ulam solver implementation, the
parallel implementation of the MCSA method will be trivial. Recall the
MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
vector operations as utilized in conventional Krylov methods. We
utilize these here for the parallel MCSA implementation. In the first
step, a parallel matrix-vector multiply is used to apply the split
operator to the previous iterate's solution. A parallel vector update
is then performed with the source vector to arrive at the initial
iteration guess. In the next step, the residual is computed by the
same operations where now the operator is applied to the solution
guess with a parallel matrix-vector multiply and then a parallel
vector update with the source vector is performed. Once the correction
is computed with a parallel adjoint Neumann-Ulam solve, this
correction is applied to the guess with a parallel vector update to
get the new iteration solution. Additionally, as given by
Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
will be required to check the stopping criteria: one initially to
compute the infinity norm of the source vector, and another at every
iteration to compute the infinity norm of the residual vector. For
this implementation, all of the issues that will be potentially
generated by the parallel adjoint solver implementation will manifest
themselves here as the quality of the correction will be of intense
study.

>>>>

%%---------------------------------------------------------------------------%%
\section{Load Balancing Concerns}
\label{sec:mc_load_balancing}
Although domain decomposition was shown to be efficient in a perfectly
load balanced situation in Siegel's work \citep{siegel_analysis_2012},
careful consideration must be made for situations where this is not
the case. Given the stochastic nature of the problem and lack of a
globally homogeneous domain, parallel Monte Carlo simulations are
inherently load imbalanced. Procassini and others worked to alleviate
some of the load imbalances that are generated by both particle and
spatial parallelism and are therefore applicable to the MSOD algorithm
\citep{procassini_dynamic_2005}. They chose a dynamic balancing scheme
in which the number of times a particular domain was replicated was
dependent on the amount of work in that domain (i.e. domains with a
high particle flux and therefore more particle histories to compute
require more work). In this variation, domains that require more work
will be replicated more frequently at reduced particle counts in each
replication. Furthermore, Procassini and colleagues noted that as the
simulation progressed and particles were transported throughout the
domain, the amount of replication for each domain would vary as
particle histories began to diffuse, causing some regions to have
higher work loads and some to have smaller work loads than the initial
conditions.

Consider the example in Figure~\ref{fig:procassini_example} adapted
from Procassini's work.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Example illustrating how domain decomposition can
      create load balance issues in Monte Carlo.}  \textit{A domain is
      decomposed into 4 zones on 8 processors with a point source in
      the lower left zone. As the particles diffuse from the source in
      the random walk sequence as shown in the top row, their tracks
      populate the entire domain. As given in the bottom row, as the
      global percentage of particles increases in a zone, that zone's
      replication count is increased.}}
  \label{fig:procassini_example}
\end{figure}
In this example, a geometry is decomposed into 4 domains on 8
processors with a point source in the bottom left domain. To begin,
because the point source is concentrated in one domain, that domain is
replicated 5 times such the amount of work it has to do per processor
is roughly balanced with the others. As the particles begin to diffuse
away from the point source, the amount of replication is adjusted to
maintain load balance. Near the end of the simulation, the diffusion
of particles is enough that all domains have equal replication.  By
doing this, load balance is improved as each domain has approximately
equal work although each domain may represent a different spatial
location and have a differing number of histories to
compute. Compared to Wagner's work where the fission source was
distributed relatively evenly throughout the domain, fixed source
problems (and especially those that have a point-like source) like
those presented in Procassini's work will be more prone to changing
load balance requirements.

%%---------------------------------------------------------------------------%%
\section{Reproducible Domain Decomposed Results}
\label{sec:reproducible_mc}
The 2006 work of Brunner is notable in that the Monte Carlo codes used
to implement and test the algorithms adhered to a strict policy of
generating identical results independent of domain decomposition or
domain replication as derived from the work of Gentile and colleagues
\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
given for obtaining results reproducible to machine precision for an
arbitrary number of processors and domains. Differences can arise from
using a different random number sequence in each domain and performing
a sequence of floating point operations on identical data in a
different order, leading to variations in round-off error and
ultimately a non-identical answer. They use a simple example,
recreated below in Figure~\ref{fig:gentile_example}, that illustrates
these issues.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/gentile_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Gentile's example illustrating how domain
      decomposition can create reproducibility issues in Monte Carlo.}
    \textit{Both particles A and B start in zone 1 on processor
      1. Particle A moves to zone 2 on processor 2 and scatters back
      to zone 1 while B scatters in zone 1 and remains there. A1 and
      A2 denote the track of particle A that is in zone 1 while B1 and
      B2 denote the track of particle B that is in zone 1.}}
  \label{fig:gentile_example}
\end{figure}
In this example, the domain is decomposed on two processors with each
processor owning one of the two zones. Starting with particle A, it is
born in zone 1 and is transported to zone 2 where a scattering event
occurs. Concerning the first reproducibility issue, if the same
sequence of random numbers is not used to compute the trajectory from
the new scattering event, we cannot expect to achieve the same result
if the domain were not decomposed. If the numbers are different, the
scattering event in zone 2 may keep the particle there or even eject
it from the domain if the sequence were different. The second issue is
demonstrated by adding another particle B that remains in the
domain. In this case, an efficient algorithm will transport particle A
on processor 1 until it leaves zone 1 and then transport particle
B. Particle A will not renter the domain until it has been
communicated to processor 2, processor 2 performs the transport, and
it is communicated back to processor 1. If we are doing a track-length
tally in zone 1, then we sum the tracks lengths observed in that
zone. In the single processor, single zone case particle A would be
transported in its entirety and then particle B transported. This
would result in a tally sum with the following order of operations:
$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
of floating point operations, we cannot expect these to have an
identical result to machine precision as round-off approximations will
differ resulting in non-commutative addition.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel Adjoint Method}
\label{sec:parallel_adjoint}
We can take much of what was learned from the survey of parallel Monte
Carlo methods for radiation transport and directly apply it to a
parallel formulation of our stochastic linear solvers. Direct analogs
can be derived from these works by noting that the primary difference
between solving a linear system with Monte Carlo methods and fixed
source Monte Carlo transport problems is the content of the Markov
chains that are generated. The transitions represented by these chains
are bound by probabilities and weights and are initiated by the
sampling of a source. In the context of transport problems, those
transitions represent events such as particle scattering and
absorption with probabilities that are determined by physical data in
the form of cross sections. For stochastic matrix inversion, those
transitions represent moving between the equations of the linear
system (and therefore the physical domain which they represent) and
their probabilities are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains. Therefore, parallel methods for Monte Carlo radiation
transport can be abstracted and we can use those concepts that apply
to matrix inversion methods as an initial means of developing a
parallel Neumann-Ulam-type solver. Based on the results observed in
the last decade of parallel Monte Carlo development, we can generate
many practical questions that this type of work could answer.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA}
\label{sec:parallel_mcsa}
With a parallel adjoint Neumann-Ulam solver implementation, the
parallel implementation of the MCSA method will be trivial. Recall the
MCSA iteration procedure outlined in Eq~(\ref{eq:mcsa}). In
\S~\ref{sec:parallel_krylov_methods} we discussed parallel matrix and
vector operations as utilized in conventional Krylov methods. We
utilize these here for the parallel MCSA implementation. In the first
step, a parallel matrix-vector multiply is used to apply the split
operator to the previous iterate's solution. A parallel vector update
is then performed with the source vector to arrive at the initial
iteration guess. In the next step, the residual is computed by the
same operations where now the operator is applied to the solution
guess with a parallel matrix-vector multiply and then a parallel
vector update with the source vector is performed. Once the correction
is computed with a parallel adjoint Neumann-Ulam solve, this
correction is applied to the guess with a parallel vector update to
get the new iteration solution. Additionally, as given by
Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
will be required to check the stopping criteria: one initially to
compute the infinity norm of the source vector, and another at every
iteration to compute the infinity norm of the residual vector. For
this implementation, all of the issues that will be potentially
generated by the parallel adjoint solver implementation will manifest
themselves here as the quality of the correction will be of intense
study.

### 2013-03-24 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
updating disseration chapter structure

\chapter{Monte Carlo Solution Methods for the Simplified $P_N$ Equations}
\label{ch:mc_spn_solutions}

@@

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis of the $SP_N$ Equations}
\label{sec:spectral_analysis}
Now that we have an understanding of the linear system generated by
the $SP_N$ approximation to the neutron transport equation and the
various parameters in the system that may be adjusted, we can consider
various means of solution. As the solution is asymmetric, modern
iterative solvers should be considered for this task. The performance
of Krylov subspace methods are bound to various properties of the
eigenvalue spectrum of the linear operator while methods based on
stationary iterations have performance limits imposed by the spectral
radius of the sytem. In this section we will review the important
spectral properties to study for common iterative methods for
asymmetric systems. Based on these properties, we will then perform a
parameter-based spectral analysis for the linear operator generated by
the multigroup $SP_N$ equations. Preconditioning strategies for these
methods are not considered. Krylov subspace method restarts and
truncation will also not be considered.

\subsection{Stationary Methods for Linear Systems}
\label{subsec:stationary_solvers}
Stationary methods for linear systems arise from splitting the
operator in Eq~(\ref{eq:multigroup_matrix_system})
\cite{leveque_finite_2007}:
\begin{equation}
  \mathbb{L} = \mathbf{M} - \mathbf{N}\:,
  \label{eq:split_linear_operator}
\end{equation}
where the choice of $\mathbf{M}$ and $\mathbf{N}$ will be dictated by
the particular method chosen. Using this split definition of the
operator we can then write:
\begin{equation}
  \mathbf{M}\mathbb{U} - \mathbf{N}\mathbb{U} = \mathbb{Q}\:.
  \label{eq:linear_split_equation1}
\end{equation}
By rearranging, we can generate a form more useful for analysis:
\begin{equation}
  \mathbb{U} = \mathbf{H}\mathbb{U} + \mathbf{c}\:,
  \label{eq:linear_split_equation2}
\end{equation}
where $\mathbf{H}=\mathbf{M}^{-1}\mathbf{N}$ is defined as the
\textit{iteration matrix} and
$\mathbf{c}=\mathbf{M}^{-1}\mathbb{Q}$. With the solution vector on
both the left and right hand sides, an iterative method can then be
formed:
\begin{equation}
    \mathbb{U}^{k+1} = \mathbf{H}\mathbb{U}^k + \mathbf{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbf{Z}^+$ defined as the \textit{iteration
  index}. Defining $\mathbf{e}^k = \mathbf{u}^k - \mathbf{u}$ as the
solution error at the $k^{th}$ iterate, we can subtract
Eq~(\ref{eq:linear_split_equation2}) from
Eq~(\ref{eq:linear_iterative_method}):
\begin{equation}
  \mathbf{e}^{k+1} = \mathbf{H}\mathbf{e}^k\:.
  \label{eq:linear_iterative_error}
\end{equation}
Our error after $k$ iterations is then:
\begin{equation}
  \mathbf{e}^{k} = \mathbf{H}^k\mathbf{e}^0\:.
  \label{eq:linear_k_iter_error}
\end{equation}
In other words, successive application of the iteration matrix is the
mechanism driving down the error in a stationary method. By assuming
$\mathbf{H}$ is diagonalizable \cite{saad_iterative_2003}, we then
have:
\begin{equation}
  \mathbf{e}^{k} =
  \mathbf{R}\boldsymbol{\Lambda}^k\mathbf{R}^{-1}\mathbf{e}^0\:,
  \label{eq:linear_k_iter_error_diag}
\end{equation}
where $\boldsymbol{\Lambda}$ contains the Eigenvalues of $\mathbf{H}$
on its diagonal and the columns of $\mathbf{R}$ contain the
Eigenvectors of $\mathbf{H}$. Computing the 2-norm of the above form
then gives:
\begin{equation}
  ||\mathbf{e}^{k}||_2 \leq ||\boldsymbol{\Lambda}^k||_2\ 
  ||\mathbf{R}||_2\ ||\mathbf{R}^{-1}||_2\ ||\mathbf{e}^0||_2\:,
  \label{eq:linear_k_iter_norm1}
\end{equation}
which gives:
\begin{equation}
  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k \kappa(\mathbf{R})
  ||\mathbf{e}^0||_2\:.
  \label{eq:linear_k_iter_norm2}
\end{equation}
For iteration matrices where the Eigenvectors are orthogonal,
$\kappa(\mathbf{R})=1$ and the error bound reduces to:
\begin{equation}
  ||\mathbf{e}^{k}||_2 \leq \rho(\mathbf{H})^k
  ||\mathbf{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}
We can now restrict $\mathbf{H}$ by asserting that $\rho(\mathbf{H}) <
1$ for a stationary method to converge such that $k$ applications of
the iteration matrix will not cause the error to grow in
Eq~(\ref{eq:linear_k_iter_norm3}). In addition, the performance of the
stationary method is dictated by $\rho(\mathbf{H})$ such that
performance of the method decreases as it approaches to unity.

\subsection{Krylov Subspace Methods for Linear Systems}
\label{subsec:subspace_solvers}
The performance of different Krylov subspace solvers can depend on
different properties of the linear system. As a means to study these
differences, we use a 1992 paper by Nachtigal and colleagues
\cite{nachtigal_how_1992} that performs and in-depth analysis of three
Krylov subspace methods that are still used today in modern physics
applications: conjugate gradient iteration for the normal equations
(CGN), generalized minimum residual iteration (GMRES), and a
biconjugate gradient-based method (CGS). For our work, we will
consider only their analysis of GMRES as it currently leveraged in
Denovo for $SP_N$ solutions and as it also applies to CGS for most
matrices.

For GMRES, at each iteration we are extracting a correction for the
solution from the Krylov subspace:
\begin{equation}
  \mathbb{U}^k \in \mathbb{U}^0 + \langle \mathbf{r_0}, \mathbb{L}
  \mathbf{r_0}, \cdots, \mathbb{L}^{k-1} \mathbf{r_0} \rangle\:,
  \label{eq:gmres_iteration}
\end{equation}
where the extraction is constrained by ensuring orthogonality of the new
residual with the Krylov space:
\begin{equation}
  \mathbf{r}^k \bot \langle \mathbb{L} \mathbf{r_0}, \mathbb{L}^2
  \mathbf{r_0}, \cdots, \mathbb{L}^{k} \mathbf{r_0} \rangle\:,
  \label{eq:gmres_constraint}
\end{equation}
with the residual at the $k^{th}$ iteration defined as:
\begin{equation}
  \mathbf{r}^k = \mathbb{Q} - \mathbb{L}\mathbb{U}^k\:.
  \label{eq:residual}
\end{equation}
From \cite{nachtigal_how_1992}, we then have the following error at
each iteration:
\begin{equation}
  \mathbf{e}^k = p_k(\mathbb{L})\mathbf{e}^0\:,
  \label{eq:gmres_iteration_error}
\end{equation}
where $p_k(\mathbb{L})$ is an arbitrary $k^{th}$ degree polynomial and
$p_k(0)=1$. This is a very similar form for the iteration error as
observed for stationary methods in Eq~(\ref{eq:linear_k_iter_norm3})
where now as these $p_k$ polynomials are constructed as the iterations
progress, a rapid decrease in $||p_k(\mathbb{L})||$ means a rapid rate
of convergence. For matrices $\mathbb{L}$ that are approximately
normal (which was exactly the assumption we made to arrive at
Eq~(\ref{eq:linear_k_iter_norm3})), then we can write the polynomial
norm as:
\begin{equation}
  ||p_k(\mathbb{L})|| = \sup_{z\in\Lambda(\mathbb{L})}|p_k(z)|\:,
  \label{eq:gmres_polynomial}
\end{equation}
where $\Lambda(\mathbb{L})$ is the eigenvalue spectrum of the normal
matrix $\mathbb{L}$. So if we could obtain the polynomials, the
largest eigenvalue of $\mathbb{L}$ could be used to calculate
convergence rates. Next we consider a choice of polynomials to use for
the convergence rate approximation. They should of course be
orthogonal to reflect the orthogonal projection mechanism by which the
residual is reduced. In Nachtigal's work, Chebyshev polynomials were
used and defined on the interval of Eigenvalues,
$[\lambda_{min},\lambda_{max}]$, and then normalized via the
constraint $p_k(0)=1$. Thus the rate of convergence for these
polynomials is an approximation to the rate of convergence of the
method.

We can check the validity of the normal
approximation by computing the condition number of $\mathbb{L}$:
\begin{equation}
  \kappa(\mathbb{L}) = ||\mathbb{L}||\ ||\mathbb{L}^{-1}||
  \label{eq:operator_condition_number}
\end{equation}
The closer $\kappa(\mathbb{L})$ is to one, the more normal the matrix
and the better the above approximation. The same check can also be
performed to verify the assumption used to generate
Eq~(\ref{eq:linear_k_iter_norm3}) where $\kappa(\mathbf{H})$ can be
computed. For our analysis, we will compute $\kappa_2$ where the
matrix norms are induced by the 2-norm of a vector such that:
\begin{equation}
  ||\mathbb{L}||_2 = \sqrt{\rho(\mathbb{L}^T\mathbb{L})}\:.
  \label{eq:matrix_2_norm}
\end{equation}

\subsection{Organization of the Spectral Analysis}
\label{subsec:spectral_organization}
For the spectral analysis we will perform a numerical parameter study
for the multigroup $SP_N$ equations. Based on the above information
for iterative methods, several key quantities should be computed for
the $SP_N$ system:
\begin{itemize}
\item Eigenvalue spectrum for the linear operator $\mathbb{L}$
\item Largest eigenvalue for the linear operator $\mathbb{L}$
\item Smallest eigenvalue for the linear operator $\mathbb{L}$
\item Condition number for the linear operator $\mathbb{L}$
\item For each stationary method:
  \begin{itemize}
  \item Eigenvalue spectrum for the iteration matrix $\mathbf{H}$
  \item Largest eigenvalue for the iteration matrix $\mathbf{H}$
  \item Smallest eigenvalue for the iteration matrix $\mathbf{H}$
  \item Condition number for the iteration matrix $\mathbf{H}$
  \end{itemize}
\item For GMRES:
  \begin{itemize}
  \item Chebyshev polynomial norm using eigenvalues of $\mathbb{L}$
    for a fixed number of iterations determined by stationary method
    convergence
  \end{itemize}
\end{itemize}
In addition, the following parameters may be varied in the $SP_N$
system in Denovo:
\begin{itemize}
  \item $SP_N$ (angular flux expansion) order
  \item $P_N$ (scattering cross section expansion) order
  \item Number of energy groups $N_g$
  \item Total cross sections $\sigma^g$
  \item Scattering cross section matrix $\sigma_{sn}^{gg'}$
  \item Boundary conditions (Marshak or reflecting)
  \item Finite volume mesh size ($\Delta$ in all directions)
  \item Number of mesh elements
  \item Source of neutrons (Uniform or Point)
\end{itemize}
We will restrict the problem description in Denovo to one similar to
Brantley and Larsen's $SP_N$ work where the spectral radius was
computed for convergence analysis of several proposed solution
methods. From this definition we choose a single material problem with
a uniform source of unity in all energy groups. For the boundary
conditions, we choose reflecting conditions on all six sides of the
cubic domain for an effectively homogenous infinite medium problem
which we will mesh with a $5\times 5\times 5$ grid with
$\Delta=0.1$. We are then left to vary the angular and energy
components of the system in our parameter studies: the $SP_N$ order,
the number of energy groups, and their cross sections of order
$P_N$. In Brantley and Larsen's work, a single energy group was used
and therefore we may not use their cross sections in our multigroup
computations. For our work, we will always choose a total cross
section of unity for all groups and all flux moments. For the
scattering matrix, two classes will be chosen: one with upscatter and
one without. For the downscatter only case, all groups may scatter
within the group and to all groups of lower energy (yielding a lower
triangular scattering matrix) with a cross section of 0.5. For the
upscatter case, all groups may scatter within the group and to all
groups of lower energy given a cross section of 0.5 and for upscatter,
all groups may scatter up to two groups higher in energy with a cross
section of 0.25. All cross section matrices will be equivalent for
each scattering moment. For these cross sections, the $P_N$ order will
be varied with $N=0,1,3$. Both 1 and 10 energy groups will be used
with the 10 energy group case using the scattering matrices specified
above. For the angular flux, the $SP_N$ order will be varied with
$N=1,3,5,7$. This gives a total of 36 linear system variations for the
spectral analysis.

\subsection{Spectral Analysis Results}
\label{subsec:analysis_results}

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration for the $SP_N$ Equations}
\label{sec:monte_carlo}

>>>>

\chapter{The Simplified $P_N$ Equations}
\label{ch:spn_equations}

@@
